\documentclass[a4paper,12pt]{article}

%\usepackage{times}
%\usepackage[T3,T1]{fontenc}
%\usepackage[utf8x]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{mathptmx}  

\usepackage{fontspec} %,xunicode} % xltxtra
    \defaultfontfeatures{Ligatures=TeX}
    \setmainfont{Brill}[
    Extension=.ttf,
    UprightFont=*-Roman,
    BoldFont=*-Bold,
    ItalicFont=*-Italic,
    BoldItalicFont=*-Bold-Italic]
    % Renderer=ICU]
% 	\usefonttheme{serif}
%	\setmainfont[Renderer=ICU]{Charis SIL}
% 	\setmainfont[Renderer=ICU]{Brill}

%% Trigger answers and notes. Package by Byron Ahn with edits by Craig Sailor
\usepackage[spaces]{myProbSol} %   Optional arguments:
%                                               'solution': reveals solutions
%                                               'spaces': leaves whitespace corresponding to the size of the solutions (has no effect if you also pass 'solution')


\usepackage{amsmath,amssymb} % for $\text{}$
\usepackage{url,natbib} 
\usepackage[dvipsnames]{xcolor} %,bm}
\usepackage{geometry,vmargin,setspace}
\usepackage{multirow}
\setmarginsrb{1in}{1in}{1in}{1in}{13.6pt}{0.1in}{0.1in}{0.2in}
% pdflscape} %rotating
    \setlength{\parindent}{0pt}

\usepackage{stmaryrd}
\usepackage{wasysym} %checkbox
\usepackage[normalem]{ulem} % for \sout{}
% \usepackage{mdwlist} % for \begin{itemize*} - but prefer \tightlist
\providecommand{\tightlist}{%
	\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{natbib}
	\bibpunct[:]{(}{)}{;}{a}{}{,}
	\setlength{\bibsep}{0pt plus 0.3ex}

\usepackage{longtable}
\usepackage[linkcolor=purple,citecolor=ForestGreen,colorlinks=true,urlcolor=gray,pagebackref=false]{hyperref}
\usepackage{multicol}
\usepackage{dashrule} %\hdashrule
\usepackage{array} % >{\...} in tabulars
\usepackage{arydshln}

\usepackage[framemethod=tikz,footnoteinside=false]{mdframed} 

\usepackage{expex} %Linguistics examples
\lingset{aboveexskip=0ex,belowexskip=0.5ex,aboveglftskip=-0.3em,interpartskip=0ex,labelwidth=!6pt,belowpreambleskip=0.1ex} %, *=*?}
%	\lingset{aboveexskip=0.5ex,belowexskip=0.5ex,*=??}

\usepackage[nocenter]{qtree} % trees
\usepackage{tikz}
    \tikzstyle{every picture}+=[remember picture]

\usepackage{tipa} % convenient for \textsubarch etc.

\newcommand\trace{\rule[-0.5ex]{0.5cm}{.4pt}}
\bibpunct[:]{(}{)}{;}{a}{}{,}
	\setlength{\bibsep}{0pt plus 0.3ex}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}% 52
\newcommand{\xmark}{\ding{55}}%
 \newcommand{\hand}{\ding{43}}

\newcommand\zero{\O{}}
\newcommand\itp[1]{\textit{\textipa{#1}}}
\newcommand\gsc[1]{\textsc{\lowercase{#1}}} %for glossing in small caps - comment out to return to caps
\renewcommand\root[1]{$\sqrt{\text{#1}}$}

\newcommand\blue[1]{\textcolor{blue}{#1}}
\newcommand\red[1]{\textcolor{red}{#1}}
\newcommand\green[1]{\textcolor{ForestGreen}{#1}}
\newcommand\gray[1]{\textcolor{gray}{#1}}
\newcommand\denote[1]{$\llbracket$#1$\rrbracket$}
\newcommand\lra{$\leftrightarrow$}

\newcommand\vz{\text{Voice$_{\text{\{--D\}}}$}}
\newcommand\vd{\text{Voice$_{\text{\{+D\}}}$}}
\newcommand\pz{\text{$p_{\text{\zero}}$}}
\newcommand\va{\root{\gsc{ACTION}}}
\newcommand{\tkal}{\emph{XaYaZ}}
\newcommand{\tpie}{\emph{XiY̯eZ}}
\newcommand{\tpua}{\emph{XuY̯aZ}}
\newcommand{\thif}{\emph{heXYiZ}}
\newcommand{\thuf}{\emph{huXYaZ}}
\newcommand{\thit}{\emph{hitXaY̯eZ}}
\newcommand{\tnif}{\emph{niXYaZ}}
\newcommand\dgs[1]{\textsubarch{#1}}
\newcommand\del[1]{\sout{#1}}

\makeatletter %Allow superscript ^ and subscript _
\catcode`_=\active%
\gdef_#1{\ensuremath{{}\sb{#1}}}%
\catcode`^=\active%
\gdef^#1{\ensuremath{{}\sp{#1}}}%
\makeatother

\begin{document}
% \pagenumbering{gobble}
% \singlespacing


\hfill \emph{Morphology 2024-25, Edinburgh, Handout 1}
\bigskip

\section{Affix ordering}

    \subsection{Auxiliaries and preliminaries}

[Exercise 1]

What's the order of auxiliaries in English?

\pex
    \a She will have be-en winn-ing the race.
    \a \ljudge{*} She has been will winning the race.
\xe
\pex
    \a The cake will have be-en (be-ing) eat-en.
    \a \ljudge{*} The cake is having were been eat.
\xe

\begin{answer}
{T > Perf > Prog > Pass > V.}
\end{answer}

\bigskip
How each element fits with the others:

\begin{itemize} \tightlist
    \item \fillblank{Takes semantic \emph{scope} over the next: the future of the perfect of the passive of an eating event.}
    \item \fillblank{Takes morpho(-phono)logical \emph{scope} over the next:}
        % \begin{itemize} \tightlist
        %     \item \fillblank{\emph{have}_{\gsc{perf}} triggers participial morphology on \emph{eat-en}}
        %     \item \fillblank{\emph{is}_{\gsc{prog}} triggers progressive morphology on \emph{winn-ing}, etc.}
        % \end{itemize}
\end{itemize}

\bigskip
Representing this formally:

\ex
\begin{answer}
{
\Tree
[.TP
    [.\emph{The cake} ]
    [.
        [.T\\\emph{will} ]
        [.PerfP/AspP
            [.Perf\\\emph{have} ]
            [.ProgP
                [.Prog\\\emph{been} ]
                [.PassP
                    [.Pass\\\emph{being} ]
                    [.VP\\\emph{eaten} ]
                ]
            ]
        ]
    ]
]
}
\end{answer}
\xe

What's the order in Latin \citep{embick10,kastnerzu17,kastner18nllt}?
\pex
    \a \begingl
        \gla am-\=a-ve-ra-m//
        \glb \root{LOVE}-\gsc{THEME}-Perf-Past-1\gsc{SG}//
        \glft `I had loved'//
        \endgl
    \a \begingl
        \gla am-\=a-ve-r-\=o//
        \glb \root{LOVE}-\gsc{THEME}-Perf-Fut-1SG//
        \glft `I will have loved'//
        \endgl
\xe

\begin{answer}
{V < Th < Perf < T.
\begin{itemize} \tightlist
    \item $\approx$ T > Perf > V.
    \item Like in English, except in morphology rather than syntax.
    \item Still get morphological conditioning, but we'll return to that later (allomorphy).
\end{itemize}}
\end{answer}

How does this compare to the order in English?
\begin{itemize} \tightlist
    \item It's the opposite \fillblank{linearly}.
    \item It's the same in terms of \fillblank{scope (hierarchically)}.
\end{itemize}

\paragraph{Summary} \begin{answer}
{Auxiliaries have a rigid ordering which reflects scope, and the fact that these auxiliaries might be independent syntactic/phonological words in one language but affixes in another.}
\end{answer}

    \subsection{The Mirror Principle}
        \subsubsection{Preliminaries}
What's the order of the reciprocal and causative suffixes in Chiche\^{w}a (spoken in Malawi, \citealt{alsina99})?
\ex[exno=3] \begingl
    \gla Al\=enje a-na-mény-\textbf{án}-\uline{its}-á mb\^{u}zi//
    \glb 2.hunters 2\gsc{s}-\gsc{PAST}-hit-\gsc{RECIP}-\gsc{CAUS}-\gsc{FinVwl} 10.goats//
    \glft `The hunters made the goats hit each other.'//
    \endgl
\xe
\ex[exno=4] \begingl
    \gla Al\=enje a-na-mény-\uline{éts}-\textbf{an}-a mb\^{u}zi//
    \glb 2.hunters \gsc{2S}-\gsc{PAST}-hit-\gsc{CAUS}-\gsc{RECIP}-\gsc{FV} 10.goats//
    \glft `The hunters made each other hit the goats.'//
    \endgl
\xe

\begin{answer}
{The order is whatever it needs to be to get scope right; you go from the verbal root ``outwards''.}
\end{answer}

\bigskip
Here's another pair \citep{hymanmchombo92}:
\pex
    \a mang-\textbf{an}-\uline{its}\\
    tie-\gsc{RECIP}-\gsc{CAUS}\\
    `cause to tie each other'
    \a mang-\uline{its}-\textbf{an}\\
    tie-\gsc{CAUS}-\gsc{RECIP}\\
    `cause each other to tie'
\xe

What would this language (\emph{Chichewa'}) look like if it had prefixes instead of suffixes?

\pex
    \a \fillblank{its-an-mang}
    \a \fillblank{an-its-mang}
\xe
Why? \fillblank{Because hierarchy is what matters (relative ordering), rather than linear order.}

\bigskip
Another example: Bemba in \cite{baker85}, citing \cite{givon76}.
\pex[exno=49]
    \a \begingl
        \gla Naa-mon-an-ya Mwape na Mutumba//
        \glb 1s.\gsc{S}-\gsc{past}-see-\gsc{recip-caus} Mwape and Mutumba//
        \glft `I made Mwape and Mutumba see each other.'//
        \endgl
    \a \begingl
        \gla Mwape na Chilufya baa-mon-eshy-ana Mutumba//
        \glb Mwape and Chilufya 3p.\gsc{S}-see-\gsc{caus-recip} Mutumba//
        \glft `Mwape and Chilufya made each other see Mutumba.'//
        \endgl
\xe

        \subsubsection{Flexible ordering reflecting interpretation}
Yupik (spoken in Alaska, \citealt[43]{mithun99}):
\pex
    \a \begingl
        \gla ayag-ciq-\textbf{yugnarqe}-\uline{ni}-llru-u-q//
        \glb go-\gsc{FUT}-probably-claim-\gsc{PAST-INDIC.INTR}-3\gsc{SG}//
        \glft `He said he would probably go.'//
    \endgl
    \a \begingl
        \gla ayag-ciq-\uline{ni}-llru-\textbf{yugnarqe}-u-q//
        \glb go-\gsc{FUT}-claim-\gsc{PAST}-probably-\gsc{INDIC.INTR}-3\gsc{SG}//
        \glft `He probably said that he would go.'//
    \endgl
\xe

Oji-Cree (spoken in Manitoba and Ontario, \citealt{slavin05uot} in \citealt{rice11}):
\pex[exno=11]
    \a \begingl
        \gla ishkwaa-niipaa-sookihpawn//
        \glb finish-at.night-be.snowing//
        \glft \fillblank{`It stopped snowing at night.' (does not snow at night anymore)}//
    \endgl
    \a \begingl
        \gla nipaa-ishkwaa-sookihpwan//
        \glb at.night-finish-be.snowing//
        \glft \fillblank{`It stopped snowing at night.' (was snowing the whole day)}// 
    \endgl
\xe
\pex[exno=11']
    \a \begingl
        \gla kiimooci-kishahtapi-wiihsini//
        \glb secretly-fast-eat//
        \glft \fillblank{`He secretly eats fast.' (nobody knows that he eats fast)}//
    \endgl
    \a \begingl
        \gla kishahtapi-kiimooci-wiihsini//
        \glb fast-secretly-eat//
        \glft \fillblank{`He eats secretly (nobody knows that he eats) and he does it fast.'}// 
    \endgl
\xe


Here's a pair of examples from Pulaar (spoken in Senegal, \citealt{paster05}), where \gsc{rep}etitive \emph{it} means `again'. Assume that one of the following means `make someone learn' and one means `teach' -- which might be which?

\pex
    \a \begingl
        \gla o jaŋŋg-in-it-ii kam//
        \glb 3\gsc{SG} learn-\gsc{CAUS-REP-past} 1\gsc{SG}//
        \glft \fillblank{`He taught me again.' (taught me before) -- [again [cause-learn]]}//
        \endgl
    \a \begingl
        \gla o jaŋŋg-it-in-ii kam//
        \glb 3\gsc{SG} learn-\gsc{REP-CAUS-past} 1\gsc{SG}//
        \glft \fillblank{`He made me learn again.' -- [cause [again-learn]]}//
        \endgl
\xe

And here \gsc{com}prehensive \emph{id} means `all' and \gsc{sep} \emph{it} denotes the reverse of the action (so `open' + \gsc{sep} = `close').
\pex
    \a \begingl
        \gla mi udd-id-it-ii baafe ɗe fof//
        \glb 1\gsc{SG} close-\gsc{COM-SEP-past} door Det all//
        \glft `I opened [sic] all the doors (in sequence).'//
        \endgl
    \a \begingl
        \gla mi udd-it-id-ii baafe ɗe fof//
        \glb 1\gsc{SG} close-\gsc{SEP-COM-past} door Det all//
        \glft `I opened all the doors (at once).'//
        \endgl
\xe

\paragraph{Summary} \begin{answer}{The order of affixes reflects the order of syntactic heads, which reflects semantic scope.\\
This is the Mirror Principle \citep{baker85}.}
\end{answer}

        \subsubsection{Extensions: Passivizing}
In Chichewa, the applicative can be used for instruments \citep{alsina99}:
\pex[exno=10] 
    \a \begingl
        \gla Ms\=odzi a-na-dúl-\textbf{ír}-a \textbf{nkhw\^{a}ngwa} uk\=onde//
        \glb 1.fisherman \gsc{1S}-\gsc{PAST}-cut-\gsc{APPL-FV} 9.axe 14.net//
        \glft `The fisherman cut the net with an axe.'//
        \endgl
\xe

Let's try to passivize: `The axe was used to cut the net (by the fisherman)'. The instrument will become the subject (we'll get back to that in argument structure). What will be the ordering of \gsc{Appl} and \gsc{Pass}?

\bigskip
\pex[exno=10]
    \a[label=b] \begingl
        \gla Nkhw\^{a}ngwa i-na-dúl-\textbf{ír}-\uline{idw}-á úk\=onde (ndí ms\=odzi)//
        \glb 9.axe 9\gsc{S}-\gsc{PAST}-cut-\gsc{APPL}-\gsc{PASS-FV} 14.net by 1.fisherman//
        \glft `The axe was used to cut the net (by the fisherman).'//
        \endgl
\xe

Schematic, ignoring the actual arguments:\\
\Tree
[.PassP
    [.Pass ]
    [.ApplP
        [.Appl ]
        \qroof{\emph{cut}}.vP
    ]
]    

\bigskip
Some more examples. Chi-Mwi:ni in \cite{baker85} from \cite{kisseberthabasheikh77}:
\pex[exno=56]   
    \a \begingl
        \gla Nu:ru Ø-chi-tes-ete chibu:ku//
        \glb Nuru \gsc{S-O}-bring-\gsc{asp} book//
        \glft `Nuru brought the book.'//
        \endgl
    \a \begingl
        \gla Nu:ru ø-m-tet-\textbf{el}-ele mwa:limu chibu:ku//
        \glb Nuru \gsc{S-O}-bring-\gsc{appl-asp} teacher book//
        \glft `Nuru brought the book to the teacher.'//
        \endgl
    \a \begingl
        \gla Mwa:limu ø-tet-\textbf{el}-el-\uline{a} chibu:ku na Nu:ru//
        \glb teacher \gsc{S}-bring-\gsc{appl-asp-pass} book by Nuru//
        \glft `The teacher was brought the book by Nuru.'//
        \endgl
\xe

Kinyarwanda in \cite{baker85} from \cite{kimenyi80}:
\pex[exno=57]
    \a \begingl
        \gla Umugabo a-ra-andik-a ibaruwa \textbf{n'i}-ikaramu//
        \glb man \gsc{S-pres}-write-\gsc{asp} letter with-pen//
        \glft `The man wrote [sic] the letter with the pen.'//
        \endgl
    \a \begingl
        \gla Umugabo a-ra-andik-\textbf{iish}-a ibaruwa ikaramu//
        \glb man \gsc{S-pres}-write-\gsc{instr-asp} letter pen//
        \glft `The man wrote the letter with the pen.'//
        \endgl
    \a \begingl
        \gla Ikaramu i-ra-andik-\textbf{iish}-\uline{w}-a ibaruwa n'umugabo//
        \glb pen \gsc{S-pres}-write-\gsc{instr-pass-asp} letter by-man//
        \glft `The pen was written-with [sic] the letter by the man.'//
        \endgl
    \a \begingl
        \gla Ibaruwa i-ra-andik-\textbf{iish}-\uline{w}-a ikaramu n'umugabo//
        \glb letter \gsc{S-pres}-write-\gsc{instr-pass-asp} pen by-man//
        \glft `The letter was written with the pen by the man.'//
        \endgl
\xe

    \subsection*{Interim summary and references}
We have seen evidence for:
\begin{itemize} \tightlist
    \item Rigid ordering for some (inflectional?) categories, e.g.~auxiliaries.
    \item Variable ordering depending on scope for some (inflectional?) categories.
    \item Affix order always respects semantic and morphological scope.
\end{itemize}

Key references:
\begin{enumerate} \tightlist
    \item \cite{baker85} coined the Mirror Principle and is the canonical work on the topic.
        \begin{itemize}
            \item \cite{muysken81,muysken88} was actually there first, at least in general linguistic publications.
            \item \cite{harley11oup} discusses different syntactic ways of deriving orders which obey or don't obey the Mirror Principle. \cite{myler17oup} analyzes a counterexample to the Mirror Principle and argues that it isn't problematic after all.
        \end{itemize}
    \item The constraints on the ordering of auxiliaries in English go back to Syntactic Structures \citep{chomsky57}. For a more contemporary introduction, see \cite{adger03}.
    \item \cite{rice11} is an excellent overview of different factors in affix ordering, based in part on \cite{rice00}. 
        % \begin{itemize}
            % \item Additional factors in affix ordering are pointed out in Section~\ref{sec:affix2}.
        % \end{itemize}
\end{enumerate}

    \subsection{Longer chains}
\pex
    \a Yupik \citep{mithun99}\\
  \begingl
    \gla ayag-yug-umi-ite-qapiar-tu-a//
    \glb go-want-be.in.state-not-really-\gsc{INDIC.INTR}-1\gsc{SG}//
    \glft `I really don't want to go.'//
    \endgl

    \a Turkish \citep[368]{inkelasorgun98} \\ \begingl
	\glpreamble \emph{çekoslovakyalilaştiramayacaklarimizdanmiydiniz}//
	\gla çekoslovakya li laş tir ama yacak lar imiz dan mi ydi niz //
	\glb Czechoslovakia from become \gsc{CAUSE} unable Fut \gsc{PL} \gsc{1PL} \gsc{ABL} \gsc{INTERR} Past \gsc{2PL}//
	\glft `were you one of those whom we are not going to be able to turn into Czechoslovakians?'//
	\endgl
\xe

Japanese:
\ex \begingl
	\gla taro-ga kodomo-o \textbf{sodat-e-sase-rare-ta}//
	\glb Taro-\gsc{NOM} child-\gsc{ACC} rise-\gsc{CAUS}-\gsc{CAUS}-\gsc{PASS}-\gsc{PAST}//
	\glft `Taro was made to raise the child.'//
	\endgl
\xe

What would a structure for this look like?

\begin{answer}
{
\ex
\Tree
[.TP
	[.\emph{taro-ga} ]
	[.
		[.vP
			[.vP
				[.Ø ]
				[.
					[.vP
						[.\sout{\emph{taro-ga}} ]
						[.
							[.VP
								[.\emph{kodomo-o} ]		
								[.V \emph{sodat} ]
							]
							[.v\\{[\gsc{caus}]} \emph{e} ]
						]
					]
					[.v\\{[\gsc{caus}]} \emph{sase} ]
				]
			]
			[.v/Voice\\{[\gsc{pass}]} \emph{rare} ]
		]
		[.T \emph{-ta} ]
	]
]
\xe
}
\end{answer}

\inSolution{Note how arguments (DPs, phrases) and suffixes live in harmony in the same tree structure.}
% \emph{sodat-e} means `raise (a child or animal)' and \emph{sodat-u} is the inchoative `grow up'

\bibliographystyle{linquiry2}
\bibliography{lingxbib}

\end{document}