\documentclass[a4paper,12pt]{article}
%\usepackage[utf8]{inputenc}
\usepackage {fontspec} %,xunicode} %,xltxtra}
%	\setmainfont{Times New Roman}
\setmainfont[
Ligatures={TeX,Common},
PunctuationSpace=0,
Numbers={Proportional},
BoldFont = LibertinusSerif-Semibold.otf,
ItalicFont = LibertinusSerif-Italic.otf,
BoldItalicFont = LibertinusSerif-SemiboldItalic.otf,
BoldSlantedFont = LibertinusSerif-Semibold.otf,
SlantedFont    = LibertinusSerif-Regular.otf,
SlantedFeatures = {FakeSlant=0.25},
BoldSlantedFeatures = {FakeSlant=0.25},
SmallCapsFeatures = {FakeSlant=0},
]{LibertinusSerif-Regular.otf}

\usepackage[margin={1in}]{geometry}

\usepackage[dvipsnames]{xcolor}
\usepackage[linkcolor=purple,citecolor=ForestGreen,colorlinks=true,urlcolor=purple,pagebackref=false]{hyperref}

\usepackage{natbib}
\bibpunct[:]{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0pt plus 0.3ex}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{expex} %Linguistics examples
\lingset{aboveexskip=0.2ex,belowexskip=0.8ex,aboveglftskip=-0.2ex,interpartskip=0.2ex,labelwidth=!6pt,belowpreambleskip=0.1ex} %, *=*?}

\usepackage{mdwlist}  %
\usepackage[normalem]{ulem}
\usepackage{array} % >{\...} in tabulars
\setlength\parindent{0pt}
\usepackage{dashrule} %\hdashrule

\usepackage{lastpage}
\usepackage{fancyhdr}
\addtolength{\headheight}{2.5pt}
\pagestyle{fancy} %or fancyplain
\lhead{Morphology 2024-25 (UoE)}
\rhead{Jigsaw group A}
\fancyfoot[C]{\thepage~of \pageref{LastPage}}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{Page \thepage~of \pageref{LastPage}} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

\definecolor{edir}{RGB}{193,0,67}
\definecolor{edib}{RGB}{0,50,95}
\newcommand\ruler{{\centering \color{edir} \rule[-0.5em]{0.6\columnwidth}{0.4pt}\par}}

\renewcommand\root[1]{$\sqrt{\text{#1}}$}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}% 52
\newcommand{\xmark}{\ding{55}}%
\newcommand{\hand}{\ding{43}}

\newcommand\gsc[1]{\textsc{\lowercase{#1}}} %for glossing in small caps - comment out to return to caps

\makeatletter %Allow superscript ^ and subscript _
\catcode`_=\active%
\gdef_#1{\ensuremath{{}\sb{#1}}}%
\catcode`^=\active%
\gdef^#1{\ensuremath{{}\sp{#1}}}%
\makeatother

\title{Jigsaw group A}
\date{Morphology 2024-25, allomorphy}
\author{}

\begin{document}

In this activity we want to understand what morpheme is \textbf{triggering} (or \textbf{conditioning}) the allomorphy of what other morpheme. The same abstract relationship holds in all cases on this sheet. For each example, we want to look at the alternations and figure out which morpheme is influencing which one; and also, on what information the alternation depends. % syntactic, outward-looking

\bigskip
The same kind of interaction is happening in~(\nextx) and also in~(\anextx). What would the \textbf{tree structure} (or bracket structure) for this interaction look like? Which node is looking at which? \textbf{What kind} of information is it looking for?

\pex
\a \emph{go} (today) 
\a  \emph{went} (yesterday) 
\xe


\pex
\a \emph{good}		 
\a  \emph{better} 
\a  \emph{best} 
\xe

\vspace{3cm}

The same phonemenon can be found for suppletive verbs in Hiaki:

\ex
\begin{tabular}{l >{\em}l >{\em}l l}
	& \gsc{SG} & \gsc{PL} & \\\hline
	a.& vuite & tenne & `run' \\
	b.& siika & saka & `go'\\
	c.& weama & rehte & `wander'\\
	d.& kivake & kiime & `enter'\\
	e.& vo'e & to'e & `lie'\\
	f.& weye & kaate & `walk'\\
	g.& mea & sua & `kill' (\gsc{sg}$\sim$\gsc{pl} object)\\
\end{tabular}
\xe

\bigskip

And the same thing happens with Applicatives in Georgian. Here the verb is \emph{rbina} `run', and the applicative \emph{i}/\emph{u} means roughly `for someone' (see if you can figure out the glosses for the different morphemes):
\pex 
\a mo-\textbf{m}-\uline{i}-rbina `\textbf{I} had run here'
\a mo-\textbf{gv}-\uline{i}-rbina `\textbf{we} had run here'
\a mo-\textbf{g}-\uline{i}-rbina(t) `\textbf{you} all had run here'\\
\hdashrule[2pt][c]{0.5\linewidth}{1pt}{1.5mm}
\a mo-\textbf{Ø}-\uline{u}-rbina(t) `\textbf{he/she/they} had run here'
\xe

%\ex \gsc{appl} \lra{} $\begin{cases}
	%	u & / [3~[~\trace{}~]] \\
	%	i
	%\end{cases}$
	%\xe

% \bigskip



\vfill

If you finish with time to spare: what other interactions would you expect to exist?


\end{document}