\documentclass[a4paper,12pt]{article}

%\usepackage{times}
%\usepackage[T3,T1]{fontenc}
%\usepackage[utf8x]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{mathptmx}  

\usepackage{fontspec} %,xunicode} % xltxtra
    \defaultfontfeatures{Ligatures=TeX}
    \setmainfont{Brill}[
    Extension=.ttf,
    UprightFont=*-Roman,
    BoldFont=*-Bold,
    ItalicFont=*-Italic,
    BoldItalicFont=*-Bold-Italic,
    Renderer=ICU]
% 	\usefonttheme{serif}
%	\setmainfont[Renderer=ICU]{Charis SIL}
% 	\setmainfont[Renderer=ICU]{Brill}

\usepackage{amsmath,amssymb} % for $\text{}$
\usepackage{url,natbib} 
\usepackage[dvipsnames]{xcolor} %,bm}
\usepackage{geometry,vmargin,setspace}
\usepackage{multirow}
\setmarginsrb{1in}{1in}{1in}{1in}{13.6pt}{0.1in}{0.1in}{0.2in}
% pdflscape} %rotating
    \setlength{\parindent}{0pt}

\usepackage{stmaryrd}
\usepackage{wasysym} %checkbox
\usepackage[normalem]{ulem} % for \sout{}
% \usepackage{mdwlist} % for \begin{itemize*} - but prefer \tightlist
\providecommand{\tightlist}{%
	\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{natbib}
	\bibpunct[:]{(}{)}{;}{a}{}{,}
	\setlength{\bibsep}{0pt plus 0.3ex}

\usepackage{longtable}
\usepackage[linkcolor=purple,citecolor=ForestGreen,colorlinks=true,urlcolor=gray,pagebackref=true]{hyperref}
\usepackage{multicol}
\usepackage{dashrule} %\hdashrule
\usepackage{array} % >{\...} in tabulars
\usepackage{arydshln}

\usepackage{expex} %Linguistics examples
\lingset{aboveexskip=0ex,belowexskip=0.5ex,aboveglftskip=-0.3em,interpartskip=0ex,labelwidth=!6pt,belowpreambleskip=0.1ex} %, *=*?}
%	\lingset{aboveexskip=0.5ex,belowexskip=0.5ex,*=??}

\usepackage[nocenter]{qtree} % trees
\usepackage{tikz}
    \tikzstyle{every picture}+=[remember picture]

\newcommand\trace{\rule[-0.5ex]{0.5cm}{.4pt}}
\bibpunct[:]{(}{)}{;}{a}{}{,}
	\setlength{\bibsep}{0pt plus 0.3ex}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}% 52
\newcommand{\xmark}{\ding{55}}%
 \newcommand{\hand}{\ding{43}}

\newcommand\zero{\O{}}
\newcommand\itp[1]{\textit{\textipa{#1}}}
\newcommand\gsc[1]{\textsc{\lowercase{#1}}} %for glossing in small caps - comment out to return to caps
\renewcommand\root[1]{$\sqrt{\text{#1}}$}

\newcommand\blue[1]{\textcolor{blue}{#1}}
\newcommand\red[1]{\textcolor{red}{#1}}
\newcommand\denote[1]{$\llbracket$#1$\rrbracket$}
\newcommand\lra{$\leftrightarrow$}

\newcommand\vz{\text{Voice$_{\text{\{--D\}}}$}}
\newcommand\vd{\text{Voice$_{\text{\{+D\}}}$}}
\newcommand\pz{\text{$p_{\text{\zero}}$}}
\newcommand\va{\root{\gsc{ACTION}}}
\newcommand{\tkal}{\emph{XaYaZ}}
\newcommand{\tpie}{\emph{XiY̯eZ}}
\newcommand{\tpua}{\emph{XuY̯aZ}}
\newcommand{\thif}{\emph{heXYiZ}}
\newcommand{\thuf}{\emph{huXYaZ}}
\newcommand{\thit}{\emph{hitXaY̯eZ}}
\newcommand{\tnif}{\emph{niXYaZ}}
\newcommand\dgs[1]{\textsubarch{#1}}
\newcommand\del[1]{\sout{#1}}

\makeatletter %Allow superscript ^ and subscript _
\catcode`_=\active%
\gdef_#1{\ensuremath{{}\sb{#1}}}%
\catcode`^=\active%
\gdef^#1{\ensuremath{{}\sp{#1}}}%
\makeatother

\begin{document}
% \pagenumbering{gobble}
% \singlespacing

\hfill \emph{EdinMorph, Nov 2021}

\section{Affix ordering}

    \subsection{Auxiliaries and preliminaries}
What's the order of auxiliaries in English?

\pex
    \a She will have be-en winn-ing the race.
    \a \ljudge{*} She has been will winning the race.
\xe
\pex
    \a The cake will have be-en (be-ing) eat-en.
    \a \ljudge{*} The cake is having were been eat.
\xe

T > Perf > Prog > Pass > V.

\bigskip
Each element:
\begin{itemize} \tightlist
    \item Takes semantic \emph{scope} over the next: the future of the perfect of the passive of an eating event.
    \item Takes morpho(-phono)logical \emph{scope} over the next: \emph{have}_{\gsc{perf}} triggers participial morphology on \emph{eat-en}, \emph{is}_{\gsc{prog}} triggers progressive morphology on \emph{winn-ing}, etc.
\end{itemize}

We've already talked about how to represent this formally:
\ex
\Tree
[.TP
    [.\emph{The cake} ]
    [.
        [.T\\\emph{will} ]
        [.PerfP/AspP
            [.Perf\\\emph{have} ]
            [.ProgP
                [.Prog\\\emph{been} ]
                [.PassP
                    [.Pass\\\emph{being} ]
                    [.VP\\\emph{eaten} ]
                ]
            ]
        ]
    ]
]
\xe

What's the order in Latin \citep{embick10,kastnerzu17,kastner18nllt}?
\pex
    \a \begingl
        \gla am-\=a-ve-ra-m//
        \glb \root{LOVE}-\gsc{THEME}-Perf-Past-1\gsc{SG}//
        \glft `I loved'//
        \endgl
    \a \begingl
        \gla am-\=a-ve-r-\=o//
        \glb \root{LOVE}-\gsc{THEME}-Perf-Fut-1SG//
        \glft `I will have loved'//
        \endgl
\xe

V < Th < Perf < T.
\begin{itemize} \tightlist
    \item $\approx$ T > Perf > V.
    \item Like in English, except in morphology rather than syntax.
    \item Still get morphological conditioning, but we'll return to that later (allomorphy).
\end{itemize}

    \subsection{The Mirror Principle}
        \subsubsection{Preliminaries}
What's the order of the reciprocal and causative suffixes in Chiche\^{w}a \citep{alsina99}?
\ex[exno=3] \begingl
    \gla Al\=enje a-na-mény-\textbf{án}-\uline{its}-á mb\^{u}zi//
    \glb 2.hunters 2\gsc{s}-\gsc{PAST}-hit-\gsc{RECIP}-\gsc{CAUS}-\gsc{FinVwl} 10.goats//
    \glft `The hunters made the goats hit each other.'//
    \endgl
\xe
\ex[exno=4] \begingl
    \gla Al\=enje a-na-mény-\uline{éts}-\textbf{an}-a mb\^{u}zi//
    \glb 2.hunters \gsc{2S}-\gsc{PAST}-hit-\gsc{CAUS}-\gsc{RECIP}-\gsc{FV} 10.goats//
    \glft `The hunters made each other hit the goats.'//
    \endgl
\xe
The order is whatever it needs to be to get scope right; You go from the verb ``outwards''.

\bigskip
Here's another pair \citep{hymanmchombo92}:
\pex
    \a mang-\textbf{an}-\uline{its}\\
    tie-\gsc{RECIP}-\gsc{CAUS}\\
    `cause to tie each other'
    \a mang-\uline{its}-\textbf{an}\\
    tie-\gsc{CAUS}-\gsc{RECIP}\\
    `cause each other to tie'
\xe

What would this language (\emph{Chichewa'}) look like if it had prefixes instead of suffixes?
\pex
    \a its-an-mang
    \a an-its-mang
\xe

Because hierarchy is what matters (relative ordering), rather than linear order.

Another example: Bemba in \cite{baker85}, citing \cite{givon76}.
\pex[exno=49]
    \a \begingl
        \gla Naa-mon-an-ya Mwape na Mutumba//
        \glb 1s.\gsc{S}-\gsc{past}-see-\gsc{recip-caus} Mwape and Mutumba//
        \glft `I made Mwape and Mutumba see each other.'//
        \endgl
    \a \begingl
        \gla Mwape na Chilufya baa-mon-eshy-ana Mutumba//
        \glb Mwape and Chilufya 3p.\gsc{S}-see-\gsc{caus-recip} Mutumba//
        \glft `Mwape and Chilufya made each other see Mutumba.'//
        \endgl
\xe

        \subsubsection{Passivizing [we skipped this]}
In Chichewa, the applicative can be used for instruments \citep{alsina99}:
\pex[exno=10] 
    \a \begingl
        \gla Ms\=odzi a-na-dúl-\textbf{ír}-a \textbf{nkhw\^{a}ngwa} uk\=onde//
        \glb 1.fisherman \gsc{1S}-\gsc{PAST}-cut-\gsc{APPL-FV} 9.axe 14.net//
        \glft `The fisherman cut the net with an axe.'//
        \endgl
\xe

Let's try to passivize: `The axe was used to cut the net (by the fisherman)'. The instrument will become the subject (we'll get back to that in argument structure). What will be the ordering of \gsc{Appl} and \gsc{Pass}?

\pex[exno=10]
    \a[label=b] \begingl
        \gla Nkhw\^{a}ngwa i-na-dúl-\textbf{ír}-\uline{idw}-á úk\=onde (ndí ms\=odzi)//
        \glb 9.axe 9\gsc{S}-\gsc{PAST}-cut-\gsc{APPL}-\gsc{PASS-FV} 14.net by 1.fisherman//
        \glft `The axe was used to cut the net (by the fisherman).'//
        \endgl
\xe

Schematic, ignoring the actual arguments:\\
\Tree
[.PassP
    [.Pass ]
    [.ApplP
        [.Appl ]
        \qroof{\emph{cut}}.vP
    ]
]    


Chi-Mwi:ni in \cite{baker85} from \cite{kisseberthabasheikh77}:
\pex[exno=56]   
    \a \begingl
        \gla Nu:ru Ø-chi-tes-ete chibu:ku//
        \glb Nuru \gsc{S-O}-bring-\gsc{asp} book//
        \glft `Nuru brought the book.'//
        \endgl
    \a \begingl
        \gla Nu:ru ø-m-tet-\textbf{el}-ele mwa:limu chibu:ku//
        \glb Nuru \gsc{S-O}-bring-\gsc{appl-asp} teacher book//
        \glft `Nuru brought the book to the teacher.'//
        \endgl
    \a \begingl
        \gla Mwa:limu ø-tet-\textbf{el}-el-\uline{a} chibu:ku na Nu:ru//
        \glb teacher \gsc{S}-bring-\gsc{appl-asp-pass} book by Nuru//
        \glft `The teacher was brought the book by Nuru.'//
        \endgl
\xe

Kinyarwanda in \cite{baker85} from \cite{kimenyi80}:
\pex[exno=57]
    \a \begingl
        \gla Umugabo a-ra-andik-a ibaruwa \textbf{n'i}-ikaramu//
        \glb man \gsc{S-pres}-write-\gsc{asp} letter with-pen//
        \glft `The man wrote [sic] the letter with the pen.'//
        \endgl
    \a \begingl
        \gla Umugabo a-ra-andik-\textbf{iish}-a ibaruwa ikaramu//
        \glb man \gsc{S-pres}-write-\gsc{instr-asp} letter pen//
        \glft `The man wrote the letter with the pen.'//
        \endgl
    \a \begingl
        \gla Ikaramu i-ra-andik-\textbf{iish}-\uline{w}-a ibaruwa n'umugabo//
        \glb pen \gsc{S-pres}-write-\gsc{instr-pass-asp} letter by-man//
        \glft `The pen was written-with [sic] the letter by the man.'//
        \endgl
    \a \begingl
        \gla Ibaruwa i-ra-andik-\textbf{iish}-\uline{w}-a ikaramu n'umugabo//
        \glb letter \gsc{S-pres}-write-\gsc{instr-pass-asp} pen by-man//
        \glft `The letter was written with the pen by the man.'//
        \endgl
\xe

        \subsubsection{Outside of Bantu}
Yupik \citep[43]{mithun99}:
\pex
    \a \begingl
        \gla ayag-ciq-\textbf{yugnarqe}-\uline{ni}-llru-u-q//
        \glb go-\gsc{FUT}-probably-claim-\gsc{PAST-INDIC.INTR}-3\gsc{SG}//
        \glft `He said he would probably go.'//
    \endgl
    \a \begingl
        \gla ayag-ciq-\uline{ni}-llru-\textbf{yugnarqe}-u-q//
        \glb go-\gsc{FUT}-claim-\gsc{PAST}-probably-\gsc{INDIC.INTR}-3\gsc{SG}//
        \glft `He probably said that he would go.'//
    \endgl
\xe
\ex \begingl
    \gla ayag-yug-umi-ite-qapiar-tu-a//
    \glb go-want-be.in.state-not-really-\gsc{INDIC.INTR}-1\gsc{SG}//
    \glft `I really don't want to go.'//
    \endgl
\xe
% (Why do we gloss these sometimes as words and sometimes as affixes? Because.)


Oji-Cree \citep{slavin05uot} in \cite{rice11}:
\pex[exno=11]
    \a \begingl
        \gla ishkwaa-niipaa-sookihpawn//
        \glb finish-at.night-be.snowing//
        \glft `It stopped snowing at night.' (does not snow at night anymore)//
    \endgl
    \a \begingl
        \gla nipaa-ishkwaa-sookihpwan//
        \glb at.night-finish-be.snowing//
        \glft `It stopped snowing at night.' (was snowing the whole day)//
    \endgl
\xe
\pex[exno=11']
    \a \begingl
        \gla kiimooci-kishahtapi-wiihsini//
        \glb secretly-fast-eat//
        \glft `He secretly eats fast.' (nobody knows that he eats fast)//
    \endgl
    \a \begingl
        \gla kishahtapi-kiimooci-wiihsini//
        \glb fast-secretly-eat//
        \glft `He eats secretly (nobody knows that he eats) and he does it fast.'//
    \endgl
\xe


Pulaar \citep{paster05}:
\pex
    \a \begingl
        \gla mi udd-id-it-ii baafe ɗe fof//
        \glb 1\gsc{SG} close-\gsc{COM-SEP-past} door Det all//
        \glft `I opened [sic] all the doors (in sequence).'//
        \endgl
    \a \begingl
        \gla mi udd-it-id-ii baafe ɗe fof//
        \glb 1\gsc{SG} close-\gsc{SEP-COM-past} door Det all//
        \glft `I opened all the doors (at once).'//
        \endgl
\xe
\pex
    \a \begingl
        \gla o jaŋŋg-in-it-ii kam//
        \glb 3\gsc{SG} learn-\gsc{CAUS-REP-past} 1\gsc{SG}//
        \glft `He taught me again.' (taught me before)//
        \endgl
    \a \begingl
        \gla o jaŋŋg-it-in-ii kam//
        \glb 3\gsc{SG} learn-\gsc{REP-CAUS-past} 1\gsc{SG}//
        \glft `‘He made me learn again.'//
        \endgl
\xe

    \subsection*{Interim summary}
\begin{itemize} \tightlist
    \item Rigid ordering for some (inflectional?) categories, e.g.~auxiliaries.
    \item Variable ordering depending on scope for some (inflectional?) categories.
    \item Always respect semantic and morphological scope.
\end{itemize}

Key references:
\cite{baker85} gets all the glory for coining the Mirror Principle but \cite{muysken81,muysken88} was there first. \cite{rice11} is an excellent overview of different factors in affix ordering, based in part on \cite{rice00}.


    \subsection{Longer chains}
Turkish \citep[368]{inkelasorgun98} 
\ex \begingl
	\glpreamble \emph{çekoslovakyalilaştiramayacaklarimizdanmiydiniz}//
	\gla çekoslovakya li laş tir ama yacak lar imiz dan mi ydi niz //
	\glb Czechoslovakia from become \gsc{CAUSE} unable Fut \gsc{PL} \gsc{1PL} \gsc{ABL} \gsc{INTERR} Past \gsc{2PL}//
	\glft `were you one of those whom we are not going to be able to turn into Czechoslovakians?'//
	\endgl
\xe

Japanese:
\ex \begingl
	\gla taro-ga kodomo-o \textbf{sodat-e-sase-rare-ta}//
	\glb Taro-\gsc{NOM} child-\gsc{ACC} rise-\gsc{CAUS}-\gsc{CAUS}-\gsc{PASS}-\gsc{PAST}//
	\glft `Taro was made to raise the child.'//
	\endgl
\xe

What would a structure for this look like?

% \ex
% \Tree
% [.IP
% 	[.\emph{taro-ga} ]
% 	[.
% 		[.vP
% 			[.vP
% 				[.Ø ]
% 				[.
% 					[.vP
% 						[.\sout{\emph{taro-ga}} ]
% 						[.
% 							[.VP
% 								[.\emph{kodomo-o} ]		
% 								[.V \emph{sodat} ]
% 							]
% 							[.v\\{[\gsc{caus}]} \emph{e} ]
% 						]
% 					]
% 					[.v\\{[\gsc{caus}]} \emph{sase} ]
% 				]
% 			]
% 			[.v\\{[\gsc{pass}]} \emph{rare} ]
% 		]
% 		[.I \emph{-ta} ]
% 	]
% ]
% \xe

% \emph{sodat-e} means `raise (a child or animal)' and \emph{sodat-u} is the inchoative `grow up'


    \subsection{Nouns}
Gender in Catalan, \cite{kramer16llc} from \cite{picallo91}:
\pex
    \a \begingl
        \gla el gos-ø//
        \glb the.\gsc{M} dog.\gsc{M}//
        \endgl
    \a \begingl
        \gla els goss-o-s//
        \glb the-\gsc{PL} dog-\gsc{M}-\gsc{PL}//
        \endgl
\xe
\pex
    \a \begingl
        \gla la goss-a//
        \glb the.F dog-F//
        \endgl
    \a \begingl
        \gla les goss-e-s//
        \glb the.\gsc{F.PL} dog-\gsc{F-PL}//
        \endgl
\xe

Schematic tree:
\ex
\Tree
[.
    [.
        [.\emph{goss} ]
        [.\gsc{[F]} \emph{e} ]
    ]
    [.\gsc{[PL]} \emph{s} ]
]
\xe

Yupik \citep[43]{mithun99}
\pex
    \a yug-\textbf{pag}-\uline{cuar}\\
    person-big-little\\
    `little giant'
    \a yug-\uline{cuar}-\textbf{pag}\\
    person-little-big\\
    `big midget'
\xe

English:
\ex glob-al-iz-ation \label{ex:globalization}
\xe
\ex novel-iz-ation-s
\xe
\ex solid-ifi-ed
\xe

Let's draw a structure for~(\ref{ex:globalization}):
\ex
\Tree
[.[N]
    [.[V]
        [.[A]
            [.[N] \emph{glob} ]
            [.[A] \emph{al} ]
        ]
        [.[V] \emph{ize} ]
    ]
    [.[N] \emph{ation} ]
]
\xe

Is this morphology or syntax? Does it matter, and if so, in what ways?

    \subsection{Building structure}
Quick reminder: what is the evidence for hierarchical structure within words, as opposed to linear structure?

        \subsubsection{Ordering}
\pex \emph{unhelpful}:
    \a {[}un-[help-ful]]
    \a \ljudge{*} {[[}un-help]-ful] 
\xe
\pex \emph{unpredictable}:
    \a {[}un-[predict-able]]
    \a \ljudge{*} {[[}un-predict]-able] 
\xe

\pex But also \emph{unhappier} (a bracketing paradox, because -\emph{er} normally attaches to smaller prosodic bases):
    \a \ljudge{*} {[}un-[happi-er]]
    \a {[[}un-happi]-er] 
\xe

See also \citep{libben03springer} for processing aspects and \cite{osekimarantz20} for modelling.

        \subsubsection{Ambiguity}
Where there's structure, there can be structural ambiguity.

\pex
    \a \emph{unlockable, unfoldable}, \dots{} \hfill \citep{dealmedialibben05,pollatseketal10}
    \a Compounds like \emph{kitchen towel rack}
\xe

        \subsubsection{A formal implementation}
Schematic trees for:
\ex  
    \Tree
    [.\emph{glob-al-iz-ation-s}
        [.[N]
            [.[V]
                [.[A]
                    [.[N] \emph{glob} ]
                    [.[A] \emph{al} ]
                ]
                [.[V] \emph{ize} ]
            ]
            [.[N] \emph{ation} ]
        ]
        [.[\gsc{PL}] \emph{s} ]
    ]
\xe

\ex 
    \Tree
    [.\emph{redd-en-ed}
        [.[V]
            [.[A] \emph{red} ]
            [.[V] \emph{en} ]
        ]
        [.[\gsc{PAST}] \emph{ed} ]
    ]
\xe

% \ex solid-ifi-ed
% \xe

Are we putting everything in the same tree or is there some point in which we want to take the ``morphological tree'' and put it within a ``syntactic tree''?

It's hard to talk about building up words (or phrases, or sentences) without a concrete theory of morphology, a concrete theory of syntax, and empirical domains in which to test them. I'm also of the opinion that it's extremely hard to talk about processing or computational aspects of morphology (or syntax) without a formal theory. So we'll flesh out one kind of theory that blurs the distinction between morphology and syntax somewhat, mostly in order to give ourselves a concrete starting point.

\bigskip
Back to technical concerns: what about bound roots?
\pex
    \a \emph{eternity} \\
    \Tree
        [.n
            [.\root{etern} ]
            [.n \emph{ity}\\{[N]} ]
        ]
    \a \emph{eternal} \\
    \Tree
        [.a
            [.\root{etern} ]
            [.a \emph{al}\\{[A]} ]
        ]
\xe

\bigskip
With that in place, what's to stop us from always positing a root? What would that look like for the noun \emph{globe}?
\ex
\Tree
[.n
    [.\root{\gsc{globe}} ]
    [.n ]
]
\xe

Nothing really, though we would need grounds to decide which of the following two derivations to prefer:
\ex \emph{global}\\
    a.~\Tree
    [.a
        [.n
            [.\root{\gsc{globe}} ]
            [.n ]
        ]
        [.a\\\emph{al} ]
    ]
    \hspace{5em}
    b.~\Tree
    [.a
        [.\root{\gsc{GLOBE}} ]
        [.a\\\emph{al} ]
    ]
\xe

\bigskip
Next, we would expect to find languages that have these kinds of roots which must be \emph{categorized} as a noun, verb or adjective first.

$\Rightarrow$ \cite{arad03}!


    \subsection{Back to inflection and derivation}

Let's recap once more the differences between inflection and derivation. We want to see whether our way of building structure helps explain these.

\begin{tabular}{ll}
Inflection & 
Derivation \\\hline
Forced by syntactic context & 
Not forced by syntactic context\\ 
Productive in all lexical categories &  
Limited productivity \\
More regular (morphologically? Semantically?) & 
Less regular \\
Does not change category &
Sometimes changes category \\ 
Does not change meaning (compositional)  &
Usually changes meaning \\
Doesn't create new stems & 
Creates new stems \\
Farther away from the base & 
Closer to the base \\
\end{tabular}

\bigskip
Starting from the last line, we now have an \emph{explanation} for why derivation is closer to the root: it's scope! It's scope all the way down!

\bigskip
We can also look at meaning change again. Do we get meaning change with \emph{every} derivational affix? Does our theory explain this?
\pex
    \a globalization
    \a novelization
\xe

\bigskip
What about non-compositional meaning?



\bibliographystyle{linquiry2}
\bibliography{lingxbib}

\end{document}
